An Analysis of Massachusetts' Standardized Testing through Multi-Group Structural Equation Modelling

### 2014-03-20 paper-knitr.Rnw
updates et cet.

Missing data in survey and test data is an important consideration, and there exist a number of techniques to produce more robust solutions, mainly working under the bayesian estimation process for missing data pioneered by Donald Rubin\cite{rubin}. However, within the structure of the exam, missing data was unsurprisingly a rare occurence: the highest proportion of missing for any item was .008, and 91.23\% had no missing questions. Even so, for the MG-CFA, we attempted to account for this by utilising the Full Information Maximum Likelihood (FIML) estimation, in conjunction with a large number of bootstraps to obtain accurate estimates of asymptotic errors. However, the questionnaire exhibited extreme proportions of non-response, which is problematic for the multivariate regression techniques, producing potentially biased results under the assumption that items are missing completely at random. As such, we processed the original responses using the mice package\cite{mice}, using 25 imputed data frames with 30 iterations in each dataframe. Regression trees are relatively robust to missing data, because all information is not processed simultaneously, but sequentially in order of variable importance, and as such we did not take any additional steps when conducting that analysis.
\section{Results}
\subsection{Exploratory factor analyses}
Given the preceeding information, we examined the Mathematics, English, and Biology sections of the exam, using a oblimin rotated factor structure, with the number of latent factors equal to the number of general domains; factor loadings were calculated using maximum likelihood estimates from the tetrachoric correlation matrices for each subset of items. The psych package was used to calculate the factor structure. Items that loaded higher than .3 on a factor were specified as indicators, and diagrams of each factor can be located in Appendix B.
\subsection{MG-CFA}
As a consequence of the unequal numbers of test takers between Biology and the rest of the science exam options, we chose to run two separate tests for measurement invariance: the first was composed of the total sample on the English and Mathematics, and the second was composed of the Biology testees ($N$=9,239); exploratory factor analyses (EFA) presented the following results:
<<fig=T>>=
fa.diagram(feng)
@

<<fig=T>>=
fa.diagram(fmat)
@

<<fig=T>>=
fa.diagram(fsci1)
@

@@

\subsubsection{Strong Invariance}
Strong Invariance does substantively change model fit above and beyond weak invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA =. , and the $\Delta$ TLI = and $\Delta$$\chi^2$=187.875. As a consequence, strict invariance is not tested, because by definition it cannot hold.
<<>>=
fitMeasures(MI.eng$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsection{Mathematics}
While the EFA presented a good model fit for a five factor model, when the SEM model imposed upon the data, it failed to converge due to underspecification. As such, factor 5 was removed, because it consisted of only four low loading items.
<<>>=
m1=~mitem4+mitem5+mitem8+mitem11+mitem12+mitem14+mitem25+mitem32+mitem35+mitem37+mitem40+mitem42
m2=~mitem6+mitem10+mitem17+mitem24+mitem39
m3=~mitem22+mitem28+mitem34+mitem38+mitem41
m4=~mitem18+mitem20+mitem21+mitem31+mitem42
mrawsc~m1+m2+m3+m4
@

@@

\subsubsection{Strong Invariance}
Strong Invariance does substantively change model fit above and beyond weak invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA =. , and the $\Delta$ TLI = and $\Delta$$\chi^2$=187.875. As a consequence, strict invariance is not tested, because by definition it cannot hold.
<<>>=
fitMeasures(MI.eng$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Strong Invariance}
Strong Invariance does substantively change model fit above and beyond weak invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA =. , and the $\Delta$ TLI = and $\Delta$$\chi^2$=187.875. As a consequence, strict invariance is not tested, because by definition it cannot hold.
<<>>=
fitMeasures(MI.eng$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsection{Questionnaire}
\subsubsection{Multivariate Regression Tree}
Without preexisting beliefs about the nature of the relationship between the response surface and the questionnaire and demographics, we chose to begin the analyses with an undirected multivariate tree regression \cite{De'ath}, conceptually similar to running a multinomial regression upon the different clusters of the response variables.

@@

These results indicate substantive effects between the future student aspirations (those who intended to attain a baccalaureate degree) and overall performance, with these students performing significantly better on all three sections. The failure for the function to further discretize the high achievers indicates that the individual differences wihin accounted for no additional information. For lower aspiring students, Ethnicity was found to be the second most important in discriminating test scores, with Blacks, Hispanics and Native Americans performing similarly worse on the Mathematics and Biology sections than the other ethnicities. The final node split discriminates between high and low scoring science students, with those who responded as rarely using scientific instruments performing worse. Cross-validated errors were high, at CVE=.905, which corresponds to $R^2$=.15
\subsubsection{Multivariate Regression}
A parametric regression model was also conducted exploring only main effects between the predictors and response. Multivariate and univariate responses were found within the tests

>>>>

Missing data in survey and test data is an important consideration, and there exist a number of techniques to produce more robust solutions, mainly working under the bayesian estimation process for missing data pioneered by Donald Rubin\cite{rubin}. However, within the structure of the exam, missing data was unsurprisingly a rare occurence: the highest proportion of missing for any item was .008, and 91.23\% had no missing questions. Even so, for the MG-CFA, we attempted to account for this by utilising the Full Information Maximum Likelihood (FIML) estimation, in conjunction with a large number of bootstraps to obtain accurate estimates of asymptotic errors. However, the questionnaire exhibited extreme proportions of non-response, which is problematic for the multivariate regression techniques, producing potentially biased results under the assumption that items are missing completely at random. As such, we processed the original responses using the mice package\cite{mice}, using 25 imputed data frames with 30 iterations in each dataframe. Regression trees are relatively robust to missing data, because all information is not processed simultaneously, but sequentially in order of variable importance, and as such we did not take any additional steps when conducting that analysis.
\section{Results}
\subsection{Exploratory factor analyses}
Given the preceeding information, we examined the Mathematics, English, and Biology sections of the exam, using a oblimin rotated factor structure, with the number of latent factors equal to the number of general domains; factor loadings were calculated using maximum likelihood estimates from the tetrachoric correlation matrices for each subset of items. The psych package was used to calculate the factor structure. Items that loaded higher than .3 on a factor were specified as indicators, and diagrams of each factor can be located in Appendix B.
\subsection{MG-CFA}
As a consequence of the unequal numbers of test takers between Biology and the rest of the science exam options, we chose to run two separate tests for measurement invariance: the first was composed of the total sample on the English and Mathematics, and the second was composed of the Biology testees ($N$=9,239); exploratory factor analyses (EFA) presented the following results:
<<fig=T>>=
setwd('~/Documents/workspace/DataCompete/upstat-competition')
fa.diagram(feng,main='English Oblimin')
@

<<fig=T>>=
fa.diagram(fmat,main='Mathematics Oblimin')
@

<<fig=T>>=
fa.diagram(fsci1,main='Biology Oblimin')
@

@@

\subsubsection{Strong Invariance}
Strong Invariance is established using the model fit indexes above and beyond weak invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA = .000 , and the $\Delta$ TLI =.000 and $\Delta$$\chi^2$=187.875.
<<>>=
fitMeasures(MI.eng$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Strict Invariance}
Strict invariance does not hold, with a significant $\Delta$ CFI = .034, $\Delta$ RMSEA = .006, $\Delta$$\chi^2$=1670.297, and the $\Delta$ TLI = .03 indicating lack of fit.
<<>>=
fitMeasures(MI.eng$fit.residuals, c('chisq','tli', 'cfi','rmsea','aic'))
@
\subsection{Mathematics}
While the EFA presented a good model fit for a five factor model, when the SEM model imposed upon the data, it failed to converge due to underspecification. As such, factor 5 was removed, because it consisted of only four low loading items.
<<>>=
m1=~mitem4+mitem5+mitem8+mitem11+mitem12+mitem14+mitem25+mitem32+mitem35+mitem37+mitem40+mitem42
m2=~mitem6+mitem10+mitem17+mitem24+mitem39
m3=~mitem22+mitem28+mitem34+mitem38+mitem41
m4=~mitem18+mitem20+mitem21+mitem31+mitem42
mrawsc~m1+m2+m3+m4
@

\subsubsection{Configural Invariance}
<<>>=
fitMeasures(MI.maths$fit.configural, c('chisq','tli', 'cfi','rmsea','aic'))
@
As the summary statistics given in the analyses show, configural invariance is supported, meaning that all four ethnicities employ the same conceptual frameworks to answer the test items. 
\subsubsection{Metric (Weak) Invariance}
Weak Invariance does impact the model fit compared to step 1, however negligably: $\Delta$ CFI = .004, $\Delta$ RMSEA = .001 , $\Delta$$\chi^2$=260.076, and the $\Delta$ TLI = .002.

<<>>=
fitMeasures(MI.maths$fit.loadings, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Strong Invariance}
Strong Invariance does not substantively change model fit above and beyond weak invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA = .000, the $\Delta$ TLI = .000 and $\Delta$$\chi^2$=167.652.

<<>>=
fitMeasures(MI.maths$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Strict Invariance}
Strong Invariance imposes a significant constraint upon model fit: $\Delta$ CFI = .017, $\Delta$ RMSEA = .005, the $\Delta$ TLI = .014 and $\Delta$$\chi^2$=998.348.

<<>>=
fitMeasures(MI.maths$fit.residuals, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Configural Invariance}
<<>>=
fitMeasures(MI.eng$fit.configural, c('chisq','tli', 'cfi','rmsea','aic'))
@
As the summary statistics given in the analyses show, configural invariance is supported, meaning that all four ethnicities employ the same conceptual frameworks to answer the test items. 
\subsubsection{Metric (Weak) Invariance}
Weak Invariance does impact the model fit compared to step 1, however negligably: $\Delta$ CFI = .002, $\Delta$ RMSEA = .000 , $\Delta$$\chi^2$=168.330, and the $\Delta$ TLI = .000.

@@

\subsubsection{Strong Invariance}
Strong Invariance does fails to substantively change model fit above and beyond metric invariance: $\Delta$ CFI = .002, $\Delta$ RMSEA = .000 , and the $\Delta$ TLI = .000 and $\Delta$$\chi^2$=187.875. 
<<>>=
fitMeasures(MI.eng$fit.intercepts, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsubsection{Strict Invariance}
Strict Invariance does substantively change model fit above and beyond strong invariance: $\Delta$ CFI = .034, $\Delta$ RMSEA = .006 , and the $\Delta$ TLI = .03 and $\Delta$$\chi^2$=1670.297. 
<<>>=
fitMeasures(MI.eng$fit.residuals, c('chisq','tli', 'cfi','rmsea','aic'))
@

\subsection{Questionnaire}
\subsubsection{Multivariate Regression Tree}
Without preexisting beliefs about the nature of the relationship between the response surface and the questionnaire and demographics, we chose to begin the analyses with an undirected multivariate tree regression \cite{De'ath}, conceptually similar to running a multinomial regression upon the different clusters analysis groups of the response variables. The tree also tests by default the possibility that which science test a student takes interacts with other outcomes --in addition, forcibly partitioning the data does not meaningfully increase model fit. We also made the decision to leave non-response as a legitimate factor level: this allows us to incorporate nonresponse as a partitioning criterion.

@@

These results indicate substantive effects between the future student aspirations (those who intended to attain a baccalaureate degree) and overall performance, with these students performing significantly better on all three sections. The failure for the function to further discretize the low achievers indicates that the individual differences wihin accounted for no additional information. For high aspiring students, ethnicity was found to be the second most important in discriminating test scores, with Blacks, Hispanics and Native Americans performing similarly worse on the Mathematics and Biology sections than the other ethnicities. The final node split discriminates between high and low scoring science students, with those who responded as rarely using scientific instruments performing worse. Cross-validated errors were high, at CVE=.905, which corresponds to $R^2$=.15
\subsubsection{Multivariate Regression}
A parametric regression model was also conducted exploring only main effects between the predictors and response. Multivariate and univariate responses were found within the tests
